Getting and Cleaning Data - Peer Assesment 1
============================================

Study Design:
-------------

Human Activity Recognition Using Smartphones Dataset
Version 1.0

Jorge L. Reyes-Ortiz, Davide Anguita, Alessandro Ghio, Luca Oneto.
Smartlab - Non Linear Complex Systems Laboratory
DITEN - Universit‡ degli Studi di Genova.
Via Opera Pia 11A, I-16145, Genoa, Italy.
activityrecognition@smartlab.ws
www.smartlab.ws

The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, we captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data. 

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain. See 'features_info.txt' for more details. 

For each record it is provided:

- Triaxial acceleration from the accelerometer (total acceleration) and the estimated body acceleration.
- Triaxial Angular velocity from the gyroscope. 
- A 561-feature vector with time and frequency domain variables. 
- Its activity label. 
- An identifier of the subject who carried out the experiment.

Code Book:
----------

The dataset includes the following files:

- 'README.txt'

- 'features_info.txt': Shows information about the variables used on the feature vector.

- 'features.txt': List of all features.

- 'activity_labels.txt': Links the class labels with their activity name.

- 'train/X_train.txt': Training set.

- 'train/y_train.txt': Training labels.

- 'test/X_test.txt': Test set.

- 'test/y_test.txt': Test labels.

The following files are available for the train and test data. Their descriptions are equivalent. 

- 'train/subject_train.txt': Each row identifies the subject who performed the activity for each window sample. Its range is from 1 to 30. 

- 'train/Inertial Signals/total_acc_x_train.txt': The acceleration signal from the smartphone accelerometer X axis in standard gravity units 'g'. Every row shows a 128 element vector. The same description applies for the 'total_acc_x_train.txt' and 'total_acc_z_train.txt' files for the Y and Z axis. 

- 'train/Inertial Signals/body_acc_x_train.txt': The body acceleration signal obtained by subtracting the gravity from the total acceleration. 

- 'train/Inertial Signals/body_gyro_x_train.txt': The angular velocity vector measured by the gyroscope for each window sample. The units are radians/second. 

- Features are normalized and bounded within [-1,1].
- Each feature vector is a row on the text file.

For more information about this dataset contact: activityrecognition@smartlab.ws


Instruction List:
-----------------

## 0: Install required packages
```{r}
install.packages("plyr")
library(plyr)
```


## 1: Merge the training and the test sets to create one data set.

Testing and training datasets are combined using the ldply package

```{r}
files <- list("./UCI HAR Dataset/test/X_test.txt", "./UCI HAR Dataset/train/X_train.txt")
data <- ldply(files, function(fn) read.table(fn)) 
```


## 2: Extract only the measurements on the mean and standard deviation for each measurement. 
Mean and standard deviations for each measurement are calculated using the apply function. 
These are combined into a single dataframe using the rbind function.
```{r}
measure_mean <- apply(data,2,mean)
measure_sd <- apply(data,2,sd)
measure_m_sd <- rbind(v_mean,v_sd)
```


## 3: Use descriptive activity names to name the activities in the data set

act_labels are used to convert the numeric actions listed in data2 to their name equivalent.

```{r}
files2 <- list("./UCI HAR Dataset/test/Y_test.txt", "./UCI HAR Dataset/train/Y_train.txt")
data2 <- ldply(files2, function(fn) read.table(fn)) 
act_labels <- read.table("./UCI HAR Dataset/activity_labels.txt")

act_names <- data2
act_names[act_names == 1] <- as.character(act_labels[1,2])
act_names[act_names == 2] <- as.character(act_labels[2,2])
act_names[act_names == 3] <- as.character(act_labels[3,2])
act_names[act_names == 4] <- as.character(act_labels[4,2])
act_names[act_names == 5] <- as.character(act_labels[5,2])
act_names[act_names == 6] <- as.character(act_labels[6,2])

data <- cbind(data,data2,act_names)
```


## 4: Appropriately labels the data set with descriptive activity names. 

Data frame names are added and tidied up.

```{r}
features <- as.vector(read.table("./UCI HAR Dataset/features.txt")[,2])
features <- c(features,"activity.label", "activity")

names(data) <- features
names(data) <- gsub("J", ".j", names(data))
names(data) <- gsub("B", ".b", names(data))
names(data) <- gsub("G", ".g", names(data))
names(data) <- gsub("M", ".m", names(data))
names(data) <- gsub("I", ".i", names(data))
names(data) <- gsub("A", ".a", names(data))
names(data) <- gsub("C", ".c", names(data))
names(data) <- gsub("-", ".", names(data))
names(data) <- gsub("X", "x", names(data))
names(data) <- gsub("Y", "y", names(data))
names(data) <- gsub("Z", "z", names(data))
```

## 5: Create a second, independent tidy data set with the average of each variable for each activity and each subject. 

Subject column is added to the dataset

```{r}
files3 <- list("./UCI HAR Dataset/test/subject_test.txt", "./UCI HAR Dataset/train/subject_train.txt")
data3 <- ldply(files3, function(fn) read.table(fn)) 
data[,ncol(data)+1] <- data3 
features <- c(features, "subject")
names(data) <- features
```

Measurement means are calculated for subject and activity splits column names are defined for both means specifying "subject" and using the activity_labels, respectively. 

```{r}
f <- factor(data$subject)
s <- split(data,f)
subject_mean <- sapply(s, function(data) colMeans(data[ , 1:(ncol(data)-3)]))
for(i in seq_along(colnames(subject_mean))){
        colnames(subject_mean)[i] <- paste0("subject_",i)        
}

f2 <- factor(data$activity)
s2 <- split(data, f2)
activity_mean <- sapply(s2, function(data) colMeans(data[ , 1:(ncol(data)-3)]))
colnames(activity_mean) <- gsub("1", "WALKING", colnames(activity_mean))
colnames(activity_mean) <- gsub("2", "WALKING_UPSTAIRS", colnames(activity_mean))
colnames(activity_mean) <- gsub("3", "WALKING_DOWNSTAIRS", colnames(activity_mean))
colnames(activity_mean) <- gsub("4", "SITTING", colnames(activity_mean))
colnames(activity_mean) <- gsub("5", "STANDING", colnames(activity_mean))
colnames(activity_mean) <- gsub("6", "LAYING", colnames(activity_mean))
```

Tidy dataset of means created by column binding the two data frames.
Written out to file called "tidy_data".

```{r}
tidy_data <- cbind(activity_mean, subject_mean)
#write.table(tidy_data, file = "tidy_data")
```
